Latent Variable Clustering for Classification of Events in the AT-TPC

### 2020-05-25 doc/Manuscript/article.tex
MPK edits Section 1

\author{R.~Solli}
\address{Department of Physics, University of Oslo, POB 1048 Oslo, N-0316 Oslo, Norway}

@@

% motivating ML work:
Because TPCs produce 3-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (akin to analysis in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently. {\color{blue} short discussion on prior work in nuclear/HEP/neutrino} In addition, some ML algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled dataset is unavailable. This could be  due to the inability to hand-label events, or the case where one must know a priori the types and behaviors of the reactions present in the detector in order to generate a simulated dataset. In the latter case, there must also exist a validation dataset of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on various ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

\section{Experimental Details} 
The goal of the experiment was to measure the evolution of the elastic and inelastic cross sections between $^{46}$Ar and protons as a function of energy (also called excitation function), and observe resonances in the composite system $^{47}$K that correspond to analog states in the nucleus $^{47}$Ar. Spectroscopic information can then be obtained from the shape and amplitude of the observed resonances \cite{Bradt2018} {\color{blue} check references: first should be Commisioning... this should be physics paper}. 
The experiment was performed at the National Superconducting Cyclotron Facility (NSCL) where a $^{46}$Ar beam was produced via fragmentation of a $^{48}$Ca beam on a $^9$Be target at about 140 MeV/u. The $^{46}$Ar isotopes were then filtered, thermalized, and finally re-accelerated to 4.6 MeV/u by a linear accelerator. This scheme was used to produce a low-emittance beam, which is necessary to guarantee a good energy resolution in the excitation function. Because the  $^{46}$Ar beam particles lose energy as they traverse the target gas volume, the position of the reaction vertex along the beam axis is directly related to the energy at which the reaction occurs. This allows the AT-TPC to measure the excitation function over a wide range of energies from a single beam energy.

The detector was placed inside the bore of a MRI solenoid energized to about 2 Tesla. This axial magnetic field served the purpose of bending the trajectories of the recoil particles in order to i) increase their length and ii) provide a measurement of their bending radius, directly related to their magnetic rigidity. Because the recoil particles travel in gas, they slow down and eventually stop, therefore their trajectories are described by 3-dimensional spirals (see \cite{Bradt2017}). One of the difficulties encountered in the analysis is that the shape of these spirals does not have any analytical form because it follows the energy-loss profile of the particles. It therefore needs to be simulated via an integration, which is numerically costly. Other difficulties are related to several experimental effects that deteriorate the quality of the data, namely saturation and cross-talk effects, as well as random noise.

The method used in \cite{Bradt2018} to analyze the data followed a 3-phase sequence: cleaning, filtering and fitting. Traditional methods were used to perform each of these tasks, and ultimately extract the scientific information, but there were severe limitations and high computational costs that would have become prohibitive had the data set been larger by an order of magnitude. In addition, \cite{Kuchera2019} demonstrated that the purity and statistics of the data would have been improved with the use of ML.
\todo{Should I give some details about how the 3 tasks were performed?}\todo{I think this is done sufficiently below}

@@

@@

We repeat this analysis using a PCA dimensionality reduction {\color{green}(please define)} on the latent space of the VGG16 model. This is done to estimate to what degree the class separating information is encoded in the entirety of the latent space, or in some select regions. The results from the PCA analysis were virtually identical to the results sans the PCA, and so we omit them for brevity.

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the K-means results or were significantly worse. Notably, the DBSCAN algorithm {\color{green} (add reference)} failed to provide any useful clustering results. We find this important as one of the significant drawbacks of K-means, and the deep clustering algorithms presented in section \ref{sec:deep_clustering}, is that they are all dependent on pre-determining the number of clusters. This is not the case for DBSCAN. 
\subsection{Autoencoder clustering}
\todo{look at grayscale images or colorbar with propto charge}

In the previous section we demonstrated a potential for clustering of events from an AT-TPC experiment. To build on this result we will in this section explore the application of the mixture of autoencoders (MIXAE) algorithm introduced in section \ref{sec:deep_clustering}. The other algorithm introduced in section \ref{sec:deep_clustering}, deep clustering with convolutional autoencoders (DCEC), consistently collapsed to just one cluster for all datasets.

In the MIXAE algorithm the hyper-parameters to adjust are all the ordinary parameters that we introduced in table \ref{tab:convae_hyperparams}. In addition to those parameters we have the weighting of the loss terms: $\theta$, $\alpha$ and $\gamma$. These weighting parameters are attached to the reconstruction loss, sample entropy and batch-wise entropy respectively. \cite{Zhang} note that these parameters are very important for the model performance and so we focus primarily on these.

\todo{but why}We empirically freeze the convolutional autoencoder hyperparameters to compress the original $128 \times 128$ to a $8 \times 8$ pixel grid using four convolutional layers. The parameters chosen for the autoencoders are listed in full in table \ref{tab:mixe_ae_hyperparams}.

\begin{table}[H]
\todo{prettify or remove}
\renewcommand*{\arraystretch}{0.5}
\centering
\caption{Hyperparameters selected for the autoencoder components of the MIXAE algorithm, see table \ref{tab:convae_hyperparams} for a full description of the parameters.}\label{tab:mixe_ae_hyperparams}
\setlength{\extrarowheight}{15pt}
\hspace*{-0.5in}
\begin{tabular}{ll}
\toprule
Hyperparameter & Value \\
\midrule
\multicolumn{2}{l}{Convolutional parameters: } \\
\midrule
Number of layers & $4$ \\
Kernels & $[3,\,3,\,3,\,3]$\\
Strides & $[2,\,2,\,2,\,2]$ \\
Filters & $[64,\, 32, \,16, \,8,]$ \\ 
\midrule
\multicolumn{2}{l}{Network parameters: } \\
\midrule
Activation & LReLu \\
Latent type & None \\
Latent dimension & 20  \\
$\beta$ & N/A \\
Batchnorm & False \\
\midrule
\multicolumn{2}{l}{Optimizer parameters: } \\
\midrule
$\eta$ & $10^{-3}$ \\
$\beta_1$ & $0.9$ \\
$\beta_2$ & $0.99$ \\
\bottomrule
\end{tabular}
\end{table}

Since there are then only three remaining hyperparameters we choose to perform a coarse grid-search as described in section \ref{sec:hyperparam_search_arch}.

>>>>

\author{R.~Solli}
\address{Expert Analytics AS, Tordenskiolds gate 6, 0160, Oslo, Norway}
\address{Department of Physics, University of Oslo, POB 1048 Oslo, N-0316 Oslo, Norway}

@@

% motivating ML work:
Because TPCs produce 3-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (akin to analysis in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently. {\color{blue} short discussion on prior work in nuclear/HEP/neutrino} In addition, some ML algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled dataset is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know a priori the types and behaviors of the reactions present in the detector in order to generate a labeled, simulated dataset. In the latter case, there must also exist a validation dataset of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

\section{Experimental Details} 
The goal of the experiment was to measure the evolution of the elastic and inelastic cross sections between $^{46}$Ar and protons as a function of energy (the {\em excitation function}), and observe resonances in the composite system $^{47}$K that correspond to analog states in the nucleus $^{47}$Ar. Spectroscopic information can then be obtained from the shape and amplitude of the observed resonances \cite{Bradt2018} {\color{blue} check references: first should be Commissioning... this should be physics paper}. 
The experiment was performed at the National Superconducting Cyclotron Facility (NSCL) where a $^{46}$Ar beam was produced via fragmentation of a $^{48}$Ca beam on a $^9$Be target at about 140 MeV/u. The $^{46}$Ar isotopes were then filtered, thermalized, and finally re-accelerated to 4.6 MeV/u by a linear accelerator. This scheme was used to produce a low-emittance beam, which is necessary to guarantee a good energy resolution in the excitation function. Because the  $^{46}$Ar beam particles lose energy as they traverse the target gas volume, the position of the reaction vertex along the beam axis is directly related to the energy at which the reaction occurs. This allows the AT-TPC to measure the excitation function over a wide range of energies from a single beam energy.

The detector was placed inside the bore of a MRI solenoid energized to $\sim 2$ Tesla. This axial magnetic field served the purpose of bending the trajectories of the recoil particles in order to i) increase their length and ii) provide a measurement of their bending radius, directly related to their magnetic rigidity. Because the recoil particles travel in gas, they slow down and eventually stop, therefore their trajectories are described by 3-dimensional spirals (see \cite{Bradt2017}). One of the difficulties encountered in the analysis is that the shape of these spirals does not have an analytical form because it follows the energy-loss profile of the particles. It therefore needs to be simulated via an integration, which is numerically costly. Other difficulties are related to several experimental effects that deteriorate the quality of the data, namely saturation and cross-talk effects, as well as random noise.

The method used in \cite{Bradt2018} to analyze the data followed a 3-phase sequence: cleaning, filtering and fitting. Traditional methods were used to perform each of these tasks, and ultimately extract the scientific information, but there were severe limitations and high computational costs that become prohibitive in data sets larger by an order of magnitude. In addition, \cite{Kuchera2019} demonstrated that the purity and statistics of the data are improved with the use of ML.

@@

{\color{orange} Yes, but also this section is a necessary bridge for reproducibility. Maybe it's better as an appendix?}

@@

We repeat this analysis using a Principle Component Analysis (PCA) dimensionality reduction \footnote{
PCA is a common techinque to find the significant variations in data by projecting the data along a subset of its covariance matrix eigenvectors \cite{Marsland2009}
}{\color{green}(please define)} {\color{orange} does this work?} on the latent space of the VGG16 model. This is done to estimate to what degree the class separating information is encoded in the entirety of the latent space, or in some select regions. The results from the PCA analysis were virtually identical to the results sans the PCA, and so we omit them for brevity.

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the K-means results or were significantly worse. Notably, the DBSCAN algorithm \cite{Ester96adensity-based}\cite{Bergstra2012} failed to provide any useful clustering results. We find this important as one of the significant drawbacks of K-means, and the deep clustering algorithm presented in section \ref{sec:mixae}, is that they are all dependent on pre-determining the number of clusters. This is not the case for DBSCAN. 
\subsection{Autoencoder clustering}
\todo{look at grayscale images or colorbar with propto charge}

In the previous section we demonstrated a potential for clustering of events from an AT-TPC experiment. To build on this result we will in this section explore the application of the mixture of autoencoders (MIXAE) algorithm introduced in section \ref{sec:mixae}.
Other algorithms for deep clustering based on autoencoder architectures were also implemented. These did not succeed in the clustering task and were omitted for brevity\footnote{The algoritms we implemented were the Deep Embedded Convolutional Clustering \cite{Guo2017} and its predecessor Deep Embedded Clustering \cite{Xie2016}}.

In the MIXAE algorithm the hyper-parameters to adjust are all the ordinary parameters that we introduced in table \ref{tab:convae_hyperparams}. In addition to those parameters we have the weighting of the loss terms: $\theta$, $\alpha$ and $\gamma$. These weighting parameters are attached to the reconstruction loss, sample entropy and batch-wise entropy respectively \cite{Zhang}. 
We focused on the tuning of the clustering hyper-parameters, and heuristically set the autoencoder hyper-parameters to a shallow $3\times3$ convolutional network.

The parameters chosen for the autoencoders are listed in full in table \ref{tab:mixe_ae_hyperparams}.

\begin{table}[H]
\todo{prettify or remove}
\renewcommand*{\arraystretch}{0.5}
\centering
\caption{Hyperparameters selected for the autoencoder components of the MIXAE algorithm, see table \ref{tab:convae_hyperparams} for a full description of the parameters.}\label{tab:mixe_ae_hyperparams}
\setlength{\extrarowheight}{15pt}
\hspace*{-0.5in}
\begin{tabular}{ll}
\toprule
Hyperparameter & Value \\
\midrule
\multicolumn{2}{l}{Convolutional parameters: } \\
\midrule
Number of layers & $4$ \\
Kernels & $[3,\,3,\,3,\,3]$\\
Strides & $[2,\,2,\,2,\,2]$ \\
Filters & $[64,\, 32, \,16, \,8,]$ \\ 
\midrule
\multicolumn{2}{l}{Network parameters: } \\
\midrule
Activation & LReLu \\
Latent dimension & 20  \\
Batchnorm & False \\
\midrule
\multicolumn{2}{l}{Optimizer parameters: } \\
\midrule
$\eta$ & $10^{-3}$ \\
$\beta_1$ & $0.9$ \\
$\beta_2$ & $0.99$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent Since there are then only three remaining hyperparameters we choose to perform a coarse grid-search as described in section \ref{sec:hyperparam_search_arch}.
